Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-05 doc/invt.tex
add a few references

Free energy calculation\cite{frenkel} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo (MC) and
molecular dynamics (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{laio2002, laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts an extended window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

Naturally, the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014, dickson2015}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
that in the previous stage).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Similarly, for metadynamics,
there are various ways of reducing the updating magnitude\cite{
  marsili2006, barducci2008, dickson2011, dama2014, dickson2015}.

@@

Consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij}
  =
  \mu_{ |i - j| }
  + \mu_{ i + j - 1 }
  + \mu_{ 2 n + 1 - i - j },
  \label{eq:w_band}
\end{equation}
%
where $\mu_0, \dots, \mu_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
\mu_0 + 2 \, \mu_1 + \cdots + 2 \, \mu_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
The matrix is translational invariant under the reflective
boundary condition\cite{bussi2006}.
%
On the right-hand side of Eq. \eqref{eq:w_band},
the first term represents the relative updating magnitude
to a neighbor that is $|i - j|$ bins away,
whereas the next two terms are added
%according to the boundary condition
to avoid unintended distortion
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006}.
%
Similarly, if $i$ represents a periodic variable,
the last two terms can be dropped, and
$|i-j|$ is evaluated
under the minimum image convention\cite{dama2014}.
This case is, however, not considered further.

@@

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
The difference is that,
in the first sampling scheme,
we choose $j$ randomly out of the $n$ bins
to emulate the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}.
%
We refer to this as the global sampling scheme.
%
In the second scheme,
we limit $j$ to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme models local sampling processes
typical in MD and straightforward MC simulations.
%
We refer to this as the local sampling scheme,

>>>>

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

@@

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts an extended window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one has to reduce the magnitude of updating
over the course of simulation.

Naturally, the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
that in the previous stage).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Similarly, for metadynamics,
there are various ways of reducing the updating magnitude,
particularly in the development of
self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.

@@

Consider the following matrix $\mathbf w$
%
\begin{equation}
  w_{ij}
  =
  \mu_{ |i - j| }
  + \mu_{ i + j - 1 }
  + \mu_{ 2 n + 1 - i - j },
  \label{eq:w_band}
\end{equation}
%
where $\mu_0, \dots, \mu_b \, (b < n)$
are a set of numbers satisfying
%
\begin{equation}
\mu_0 + 2 \, \mu_1 + \cdots + 2 \, \mu_b = 1,
\label{eq:m_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
The matrix is translational invariant under the reflective
boundary condition\cite{bussi2006}.
%
On the right-hand side of Eq. \eqref{eq:w_band},
the first term represents the relative updating magnitude
to a neighbor that is $|i - j|$ bins away,
whereas the next two terms are added
%according to the boundary condition
to avoid unintended distortion\cite{dickson2011, mcgovern2013}
to the equilibrium distribution $p_i = 1/n$\cite{bussi2006}.
%
Similarly, if $i$ represents a periodic variable,
the last two terms can be dropped, and
$|i-j|$ is evaluated
under the minimum image convention\cite{dama2014}.
This case is, however, not considered further.

@@

Since the error depends on the underlying sampling scheme,
we have applied the above protocol to two
sampling schemes.
%
In both cases, we used the
Metropolis algorithm\cite{metropolis1953, newman, frenkel, landau_binder}
to sample the state space:
%
in each step, a new bin index $j$ was proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(v_i - v_j) \}.
$
The difference is that,
in the first sampling scheme,
we choose $j$ randomly out of the $n$ bins
to emulate the perfect sampling scheme
discussed in Sec. \ref{sec:Gamma}.
%
We refer to this as the global sampling scheme.
%
In the second scheme,
we limit $j$ to the two neighbors of $i$,
$i \pm 1$, with equal probability $1/2$,
to mimic the one-step sampling scheme
in Sec. \ref{sec:Gamma}.
%
This scheme models local sampling processes
typical in MD and straightforward MC simulations.
%
We refer to this as the local sampling scheme,
